
---
title: "R Notebook"
output: html_notebook
---
```{r setup}
library(Matrix)  # sparseMatrix()
library(dplyr)
library(vsp)
library(tidytext)
library(ggplot2)
library(wordspace)
library(tidyverse)
library(maps)
library(ggthemes)
library(broom)
library(seqinr)
```


```{r}
#read file and libraries
infile = "otu_sample_value.txt"
d = read.table(infile, header=T, sep="\t", colClasses=c(rep("character", 2), "numeric"))
otus = read.table("row_metadata.txt", header=T, sep="\t", colClasses=rep("character", 9))
samples= read.table("column_metadata.txt", header=T, sep="\t", colClasses=rep("character", 2))
sample_meta.raw = read.csv("emp_qiime_mapping_qc_filtered.tsv",sep="\t")%>%rename(SAMPLE=X.SampleID)
#left_join samples df with empo_3 for subsetting usage
selected.samples<-samples%>%
  left_join(sample_meta.raw%>%
  select(c(SAMPLE,empo_3,ph,collection_timestamp,latitude_deg,longitude_deg,env_biome,env_feature,envo_biome_0,envo_biome_1,envo_biome_2,envo_biome_3,envo_biome_4,adiv_observed_otus,adiv_chao1,adiv_shannon,adiv_faith_pd)))%>%
  filter(empo_3%in%c("Soil (non-saline)","Plant rhizosphere"))%>%
  filter(!is.na(ph))%>%
  mutate(id_num=readr::parse_number(ID)+1)%>%
  mutate_if(is.factor,as.character)
selected.samples
levels(as.factor(selected.samples$envo_biome_2))
#samples_empo%>%group_by(country)%>%summarize(count=n())
#subset soil(non-saline) and Plant Rhizosphere sample sites
# selected.samples=rbind(
#   subset(samples_empo,samples_empo$empo_3=="Soil (non-saline)",ID),
#   subset(samples_empo,samples_empo$empo_3=="Plant rhizosphere",ID)
# )%>%mutate(id_num = (readr::parse_number(ID))+1)
idx.soilplant=as.numeric(selected.samples$id_num)

# extract row / column indices
I = as.integer(gsub("otu_", "", d$OTU)) + 1        # add 1 to convert 0-index to 1-index
J = as.integer(gsub("sample_", "", d$SAMPLE)) + 1  # add 1 to convert 0-index to 1-index
```
## plot sample global location
```{r}
selected.samples
```


## plot overall trend of pH v.s. observed sequence diversity(1500 means that there are 1500 different kinds of sequences observed in the sample site)
```{r}
ph_alpha<-selected.samples%>%select(ph,adiv_observed_otus)
ph_alpha_trend_overall<-ggplot(ph_alpha,aes(x=ph,y=adiv_observed_otus))+
  geom_point()+
  theme_bw()+
  xlab("pH")+
  ylab("observed sequences")
ggsave("ph_alpha_trend_overall.jpg",ph_alpha_trend_overall)
```



```{r}
# sparse bipartitie OTU-sample adjacency matrix (w/o filtering)
A = sparseMatrix(i=I, j=J, x=d$VALUE)
rownames(A) = otus$ID
colnames(A) = samples$ID
dim(A)

A.simple=(A > 0) * 1
#index of OTUs with prevalence of at least 10 samples(use the rowSum of A.simple to pick)
#since A.simple records 1 in Aij if otui appears in sample j; 0 otherwise
idx.prv = rowSums(A.simple) > 9

##here, just want to see if there's a difference aftering taking sqrt transformation
#take sqrt transformation(stablization)
#A.sqrt=A
#A.sqrt@x=sqrt(A.sqrt@x)
#plot(A@x)
#plot(A.sqrt@x)

#plot out-degree to examine
hist(log10(rowSums(A)))#highly skewed, some microbiomes are more abundant than others
```

```{r}
A.sel=A[idx.prv,idx.soilplant]
colnames(A.sel)
dim(A.sel)
hist(log10(rowSums(A.sel)))
#take square root transformation on A.sel (the adjmatrix of interest)
A.sel@x=sqrt(A.sel@x)
hist(log10(rowSums(A.sel)))
n=nrow(A.sel)
d=ncol(A.sel)
In<-as.matrix(c(rep(1,n)))
Id<-as.matrix(c(rep(1,d)))
length(A.sel@x)
A.sel_center<-A.sel
entries_len<-length(A.sel_center@x)
A.sel_center@x<-A.sel_center@x-(In%*%(t(In)%*%A.sel/n))[1:entries_len]
dim(A.sel_center)
```

```{r,message=FALSE}
#fa_center<-vsp(A.sel_center,rank = 10,center=FALSE) #this works too
#fa_center<-vsp(A.sel,rank = 6,center = TRUE,scale = TRUE,rescale = TRUE,recenter = FALSE)#takes forever to run

#envo_biome2 has 6 biome labels for soil-related samples
fa_center<-vsp(A.sel_center,rank = 6,center = FALSE)
pairplot_y<-plot_varimax_y_pairs(fa_center_ph,1:6)
ggsave(pairplot_y,filename = "pairplot_y.jpg")
#fa_center$Y
screeplot(fa_center)
bmtx<-plot_mixing_matrix(fa_center)
dim(fa_center$Z)
dim(A.sel_center)
ggsave(bmtx,filename = "bmatrix.jpg")
```

```{r}
samples_selected_metaInterests<-selected.samples%>%
  select(ID,env_biome,env_feature,ph)%>%
  mutate_all(as.character)
samples_selected_metaInterest1<-selected.samples%>%select(ID,env_biome)
samples_selected_metaInterest2<-selected.samples%>%select(ID,env_feature)

samples_selected_metaInterests_pivot<-pivot_longer(samples_selected_metaInterests,c(env_biome,env_feature,ph),"feature_type")
samples_selected_metaInterest12_pivot<-pivot_longer(samples_selected_metaInterests,c(env_biome,env_feature),"feature_type")
samples_meta_mtx<-cast_sparse(samples_selected_metaInterest1,ID,env_biome)
samples_meta_mtx2<-cast_sparse(samples_selected_metaInterest2,ID,env_feature)
samples_meta_mtxall<-cast_sparse(samples_selected_metaInterests_pivot,ID,value)
samples_meta_mtx12<-cast_sparse(samples_selected_metaInterest12_pivot,ID,value)

bffwith_biome<-bff(fa_center$Y,samples_meta_mtx,2)
bffwith_feature<-bff(fa_center$Y,samples_meta_mtx2,3)
#bff with biome, feature, ph
bffwith_bfp<-bff(fa_center$Y,samples_meta_mtxall,15)
#bff with biome,feature
bffwith_bf<-bff(fa_center$Y,samples_meta_mtx12,10)
bffwith_bf
bffwith_biome
bffwith_feature
bffwith_bfp
#contextualize with biome, feature, ph
bffwith_bfpdf<-data.frame(bffwith_bfp) #dataframe version
write.csv(bffwith_bfp,file="bffwith_bfp.csv",row.names = FALSE)
```


## calculate the percentage of biome kind within each community 
```{r}
phi=t(fa_center$Z)%*%A.sel_center
dim(phi)#6*945
phi@x

norm_phi=rowNorms(phi,method="minkowski",p=1)#calculate l1 norm for each row
norm_phi_d<-diag(norm_phi) #20*20 diagonal matrix
norm_phi_d_inv<-solve(norm_phi_d) #inverse of norm_phi_d
beta<-t(norm_phi_d_inv%*%phi) #2000*20
beta[beta<0]=0
beta<-apply(beta,2,function(x) return(x/sum(x)))
#beta@x<-abs(beta@x) #take absolute value? nahhh, this is one way, but usually what we do is make the negative values = 0, and calculate x/sum(x), 'cause by assumption, we don't have negative betas, so, if we get negatvie ones, we'll just say it's statistical error
colSums(beta)#each column sum up to 1
beta
hist(log10(beta))
#how samples compose each community
community_composition<-data.frame(beta)%>%rownames_to_column(var="ID")
```
## plot biome distribution within each community
```{r}
#left-join sample composition with feature of interest
##by biome
community_composition_bybiome<-samples_selected_metaInterest%>%
  left_join(community_composition)%>%
  select(-ID)%>%
  group_by(env_biome)%>%
  summarize_each(funs(sum))

##by feature
community_composition_byfeat<-samples_selected_metaInterest2%>%
  left_join(community_composition)%>%
  select(-ID)%>%
  group_by(env_feature)%>%
  summarize_each(funs(sum))

community_composit_tidy<-community_composition_bybiome%>%
  pivot_longer(!env_biome,names_to="cluster",values_to="ratio")%>%
  mutate(ratio=100*ratio)%>%
  mutate(env_biome=str_remove_all(env_biome,"biome"))%>%
  mutate(cluster=factor(cluster,levels=paste0("X",c(1:6))))

community_composit_tidy2<-community_composition_byfeat%>%
  pivot_longer(!env_feature,names_to="cluster",values_to="ratio")%>%
  mutate(ratio=100*ratio)%>%
  mutate(env_feature=str_remove_all(env_feature,c("feature","habitat")))%>%
  mutate(env_feature=str_replace_all(env_feature,"environmental","env."))%>%
  mutate(env_feature=str_replace_all(env_feature,"anthropogenic","anthrop."))%>%
  mutate(env_feature=str_replace_all(env_feature,"hydrothermal","hydro-thermal"))%>%
  mutate(env_feature=str_replace_all(env_feature,"cultivated","culti-vated"))%>%
  mutate(cluster=factor(cluster,levels=paste0("X",c(1:6))))
new_labels<-c("X1"="Weathered","X2"="Cold moist","X3"="Dry","X4"="Short vegetation","X5"="Cold dry","X6"="Disturbed")
p<-ggplot(community_composit_tidy[which(community_composit_tidy$ratio>9),], aes(x=env_biome,y=ratio)) +
  geom_col() +
  facet_wrap( ~ cluster, scales="free",ncol=3,labeller=labeller(cluster=new_labels)) +
  theme_bw()+
  aes(stringr::str_wrap(env_biome, 5), ratio)+xlab("biome type")+
  theme(axis.text.x=element_text(size=rel(0.8)),axis.text.y = element_text(size=rel(0.8)))

p
ggsave(p,filename = "env_biome_weight.jpg")
p2<-ggplot(community_composit_tidy2[which(community_composit_tidy2$ratio>9),], aes(x=env_feature,y=ratio)) +
  geom_col() +
  facet_wrap( ~ cluster, scales="free",ncol=3) +
  theme_bw()+
  aes(stringr::str_wrap(env_feature, 3), ratio)+xlab("env feature")+
  theme(axis.text.x=element_text(size=rel(1.0)),axis.text.y = element_text(size=rel(1.0)))
p
p2
ggsave(p2,filename = "env_feature_weight.jpg")
```
## get memberships
```{r}
sample_membership<-fa_center%>%
  get_varimax_y()%>%
  select(-id)%>%
  mutate_all(abs)%>%
  apply(1,which.max)
#add a column that says the sample's membership
selected.samples<-selected.samples%>%mutate(cluster_mem=sample_membership)
selected.samples
```
## plot pH distribution of each cluster
```{r}
sample_phbycluster<-selected.samples%>%select(ID,ph,cluster_mem)
ph_cluster<-ggplot(sample_phbycluster, aes(x=as.factor(cluster_mem), y=ph)) + 
  geom_boxplot()+
  xlab("cluster")+
  scale_x_discrete(labels=c("weathered","cold moist","dry","short vegatation","dry cold","disturbed"))+
  ylab("pH")+
  theme_bw()
ph_cluster
```


## within each cluster, pH v.s. diversity
```{r}
#use alpha diversity: phylum diversity
selected.samples
pdiv_bycluster<-selected.samples%>%select(ID,ph,cluster_mem,adiv_faith_pd)
new_labels<-c("1"="Weathered (acidic)","2"="Cold moist","3"="Dry (basic)","4"="Short vegetation (neutral)","5"="Dry cold","6"="Disturbed")
p3<-ggplot(pdiv_bycluster,aes(x = ph,y=log10(adiv_faith_pd)))+
  geom_point()+
  facet_wrap(~cluster_mem,scales="free",labeller = labeller(cluster_mem=new_labels))+
  theme_bw()+
  xlab("pH")+ylab("phyla diversity")
#fix x,y axis
p4<-ggplot(pdiv_bycluster,aes(x = ph,y=adiv_faith_pd))+
  geom_point()+
  facet_wrap(~cluster_mem,labeller = labeller(cluster_mem=new_labels))+
  theme_bw()+
  xlab("pH")+ylab("phyla diversity")
p4
ggsave("ph_phydiv_bycluster.jpg",p4)

bycluster<-pdiv_bycluster%>%group_by(cluster_mem)
do(bycluster, 
+    glance( 
+       lm(adiv_faith_pd~ph, data = .)))
fits=pdiv_bycluster%>%group_by(cluster_mem)%>%do(model=lm(adiv_faith_pd~ph,data=.))
lapply(fits$model, function(x)summary(x))
```
## community membership globally
```{r}

world <- ggplot() +
  borders("world", colour = "gray85", fill = "gray80") +
  theme_map() +
  coord_cartesian(ylim = c(-50, 90)) 

membership_id<-c(1,2,3,4,5,6)
membership_name<-c("1: Weathered (acidic)","2: Cold moist","3: Dry (basic)","4: Short vegatation (neutral)","5: Dry cold","6: Disturbed")
membership_dic<-data.frame(cluster_mem=membership_id,membership_name)
global_membership<-selected.samples%>%
  select(latitude_deg,longitude_deg,cluster_mem)%>%
  left_join(membership_dic)%>%
  select(-cluster_mem)
global_membership
map <- world +
  geom_point(aes(x = longitude_deg, y = latitude_deg, colour = membership_name),size=2,
             data = global_membership) +
  labs(colour = 'Cluster membership')
map
ggsave("membership_map.jpg",map)
```
## relate cluster to phyla/genus/function composition  
```{r}
otus
selected.otus_ID<-row.names(A.sel)
dim(A.sel)
selected.otus<-otus %>% filter(ID%in%selected.otus_ID)%>%select(ID,PHYLUM)
selected.otus2<-otus %>% filter(ID%in%selected.otus_ID)%>%select(ID,GENUS)
selected.otus3<-otus %>% filter(ID%in%selected.otus_ID)%>%select(ID)%>%left_join(taxfunc)
selected.otus3
selected.otus_mtx<-cast_sparse(selected.otus,ID,PHYLUM)
selected.otus_mtx2<-cast_sparse(selected.otus2,ID,GENUS)
selected.otus_mtx3<-cast_sparse(selected.otus3,ID,func)

dim(selected.otus_mtx)
dim(selected.otus_mtx3)

bffwith_phyla<-bff(fa_center$Z,selected.otus_mtx,10)
bffwith_genus<-bff(fa_center$Z,selected.otus_mtx2,10)
bffwith_func<-bff(fa_center$Z,selected.otus_mtx3,10)

bffwith_phyla
bffwith_genus
bffwith_func

bfffuncdf<-data.frame(bffwith_func)
write.csv(bfffuncdf,"bffwithfunction.csv",row.names = FALSE)
```

## community composition by phyla 
```{r}
phi=t(A.sel_center%*%fa_center$Y)
dim(phi)#6*945
phi@x

norm_phi=rowNorms(phi,method="minkowski",p=1)#calculate l1 norm for each row
norm_phi_d<-diag(norm_phi) #6*6 diagonal matrix
norm_phi_d_inv<-solve(norm_phi_d) #inverse of norm_phi_d
beta<-t(norm_phi_d_inv%*%phi) #6*6
beta[beta<0]=0
beta<-apply(beta,2,function(x) return(x/sum(x)))
#beta@x<-abs(beta@x) #take absolute value? nahhh, this is one way, but usually what we do is make the negative values = 0, and calculate x/sum(x), 'cause by assumption, we don't have negative betas, so, if we get negatvie ones, we'll just say it's statistical error
colSums(beta)#each column sum up to 1
beta
hist(log10(beta))
#how otus compose each community
community_composition_otu<-data.frame(beta)%>%rownames_to_column(var="ID")
```

## plot community composition by phyla(highlight the ones that are top prominant in the cluster)
```{r}
new_labels<-c("X1"="Weathered","X2"="Cold moist","X3"="Dry","X4"="Short\nvegetation","X5"="Cold dry","X6"="Disturbed")
##community composition by phyla
community_composition_byphyla<-selected.otus%>%
  left_join(community_composition_otu)%>%
  select(-ID)%>%
  group_by(PHYLUM)%>%
  summarize_each(funs(sum))
community_composition_byphyla
#add a column to highligh more related phyla for each cluster
community_composit_tidy_phyla<-community_composition_byphyla%>%
  pivot_longer(!PHYLUM,names_to="cluster",values_to="ratio")%>%
  mutate(ratio=100*ratio)%>%
  mutate(PHYLUM=str_remove_all(PHYLUM,"p__"))%>%
  mutate(cluster=factor(cluster,levels=paste0("X",c(1:6))))%>%
  mutate(highlight=ifelse(ratio>8,"yes","no"))
community_composit_tidy_phyla
#prominant phyla within a community
ppwc<-community_composit_tidy_phyla[which(community_composit_tidy_phyla$ratio>8),]$PHYLUM
ppwclevel<-levels(as.factor(ppwc))
ppwclevel

p5<-ggplot(community_composit_tidy_phyla[which(community_composit_tidy_phyla$ratio>8),], aes(x=PHYLUM,y=ratio)) +
  geom_col() +
  facet_wrap( ~ cluster, scales="free",ncol=3,labeller = labeller(cluster=new_labels)) +
  theme_bw()+
  aes(stringr::str_wrap(PHYLUM, 2), ratio)+xlab("phyla")+
  theme(axis.text.x=element_text(size=rel(0.8)),axis.text.y = element_text(size=rel(0.8)))
#adjust the plot to show same phyla in each cluster
p5.1<-ggplot(community_composit_tidy_phyla[which(community_composit_tidy_phyla$PHYLUM%in%ppwclevel),], aes(x=PHYLUM,y=ratio,fill=highlight)) +
  geom_col() +
  facet_wrap( ~ cluster,ncol=2,labeller = labeller(cluster=new_labels)) +
  theme_bw()+
  scale_fill_manual(values = c("yes"="tomato","no"="gray"),guide=FALSE)+
  aes(stringr::str_wrap(PHYLUM, 2), ratio)+xlab("phyla")+
  theme(axis.text.x=element_text(size=rel(1.5)),axis.text.y = element_text(size=rel(0.8)), strip.text.x = element_text(size = rel(1.5)))

community_composit_tidy_phyla
p5
p5.1
ggsave("phyla_composit.jpg",p5.1,width=1920/100,height=1080/100,dpi=600)
```
## plot communtiy composition by genus
```{r}
## community composition by genus
## filter out no-genus data
community_composition_bygenus<-selected.otus2%>%
  left_join(community_composition_otu)%>%
  select(-ID)%>%
  group_by(GENUS)%>%
  summarize_each(funs(sum))%>%
  filter(GENUS!="")%>%
  filter(GENUS!="g__")
community_composition_bygenus
community_composit_tidy_genus<-community_composition_bygenus%>%
  pivot_longer(!GENUS,names_to="cluster",values_to="ratio")%>%
  mutate(ratio=100*ratio)%>%
  mutate(GENUS=str_remove_all(GENUS,"g__"))%>%
  mutate(cluster=factor(cluster,levels=paste0("X",c(1:6))))
community_composit_tidy_genus

p6<-ggplot(community_composit_tidy_genus[which(community_composit_tidy_genus$ratio>1),], aes(x=GENUS,y=ratio)) +
  geom_col() +
  facet_wrap( ~ cluster, scales="free",ncol=3) +
  theme_bw()+
  aes(stringr::str_wrap(GENUS, 2), ratio)+xlab("genus")+
  theme(axis.text.x=element_text(size=rel(0.8)),axis.text.y = element_text(size=rel(0.8)))

p6
```



## map otu to phyla for each sample
```{r}
# join otu-sample with phyla
otuPhy<-otus%>%select(ID,PHYLUM)%>%mutate(PHYLUM=str_remove_all(PHYLUM,"p__"))%>%rename(OTU=ID)
otuSamplePhy<-read.table(infile, header=T, sep="\t", colClasses=c(rep("character", 2), "numeric"))%>%
  left_join(otuPhy)
otuSamplePhy
#for each sample site, sum the otu number of each phylum
#each sample site has 5000 otus
#exclude otus with unknown phylum
perphylsumbysample<-otuSamplePhy%>%
  group_by(SAMPLE,PHYLUM)%>%
  summarize(amount=sum(VALUE))%>%
  mutate(abundance=(amount/5000)*100)%>%
  filter(PHYLUM!="")
#sample-clustermembership
sample_clustermem<-selected.samples%>%select(ID,cluster_mem)%>%rename(SAMPLE=ID)
#selected samples that have memberships(we only included samples that have otu num>9, that's why some samples were not included)
perphylsumbysample_withmem<-perphylsumbysample%>%
  filter(SAMPLE%in%sample_clustermem$SAMPLE)%>%
  left_join(sample_clustermem)%>%
  rename(cluster=cluster_mem)
perphylsumbysample_withmem
#within each cluster, phyla that have more weight
more_relatedphyla<-(community_composit_tidy_phyla[which(community_composit_tidy_phyla$ratio>8),])%>%
  mutate_if(is.factor,as.character)%>%
  mutate(cluster=str_remove_all(cluster,"X"))%>%
  mutate(cluster=as.integer(cluster))
# more_relatedphyla<-(community_composit_tidy_phyla[which(community_composit_tidy_phyla$PHYLUM%in%ppwclevel),])%>%
#   mutate_if(is.factor,as.character)%>%
#   mutate(cluster=str_remove_all(cluster,"X"))%>%
#   mutate(cluster=as.integer(cluster))
more_relatedphyla
perphylsumbysample_withmem
#for each cluster, take the samples that have higher-weight-phyla in that cluster
sample_ofrelatedphyla_bycluster<-perphylsumbysample_withmem%>%
  left_join(more_relatedphyla,by=c("PHYLUM","cluster"))%>%
  #if the ratio column is na, it means that the phylum-cluster key is not in the more_relatedphyla, 
  #meaning that the phylum is not that important in the cluster; thus excluse
  filter(!is.na(ratio))%>%
  select(-ratio)
sample_ofrelatedphyla_bycluster
sample_ph<-selected.samples%>%select(ID,ph,envo_biome_2)%>%rename(SAMPLE=ID)
#combine sample_ofrelatedphyla_bycluster with sample_ph
sorpc_ph<-sample_ofrelatedphyla_bycluster%>%left_join(sample_ph)
sorpc_ph
```
## plot ph by phyla composition
```{r}
new_labels<-c("1"="Weathered","2"="Cold moist","3"="Dry","4"="Short\nvegetation","5"="Cold dry","6"="Disturbed")
#facet grid is clearer than facet wrap
p7<-ggplot(sorpc_ph,aes(x = ph,y=log(abundance),color=envo_biome_2))+
  geom_point()+
  facet_grid(cluster~PHYLUM,labeller = labeller(cluster=new_labels))+
  theme_bw()+
  xlab("pH")+ylab("log(phyla abundance)")+
  theme(strip.text.x = element_text(size = 14,colour = "black"),
        strip.text.y = element_text(size=14),
        legend.title = element_blank(),
        legend.text = element_text(size = 16),
        axis.text=element_text(size=12,colour = "black"),
        axis.title=element_text(size=15,colour = "black"),
        legend.position = "none")
p7
ggsave("ph-phyla.jpg",p7,width=1920/100,height=1080/100,dpi=600)
```
## map otu to genus for each sample
```{r}
# join otu-sample with genus
otugen<-otus%>%select(ID,GENUS)%>%mutate(GENUS=str_remove_all(GENUS,"g__"))%>%rename(OTU=ID)
otuSampleGen<-read.table(infile, header=T, sep="\t", colClasses=c(rep("character", 2), "numeric"))%>%
  left_join(otugen)
otuSampleGen
#for each sample site, sum the otu number of each genus
#each sample site has 5000 otus
#exclude otus with unknown genus
pergensumbysample<-otuSampleGen%>%
  group_by(SAMPLE,GENUS)%>%
  summarize(amount=sum(VALUE))%>%
  mutate(abundance=(amount/5000)*100)%>%
  filter(GENUS!="")
#sample-clustermembership
sample_clustermem<-selected.samples%>%select(ID,cluster_mem)%>%rename(SAMPLE=ID)
#selected samples that have memberships(we only included samples that have otu num>9, that's why some samples were not included)
pergensumbysample_withmem<-pergensumbysample%>%
  filter(SAMPLE%in%sample_clustermem$SAMPLE)%>%
  left_join(sample_clustermem)%>%
  rename(cluster=cluster_mem)
pergensumbysample_withmem
#within each cluster, phyla that have more weight
more_relatedgen<-(community_composit_tidy_genus[which(community_composit_tidy_genus$ratio>1),])%>%
  mutate_if(is.factor,as.character)%>%
  mutate(cluster=str_remove_all(cluster,"X"))%>%
  mutate(cluster=as.integer(cluster))
more_relatedgen
pergenlsumbysample_withmem
#for each cluster, take the samples that have higher-weight-phyla in that cluster
sample_ofrelatedgen_bycluster<-pergensumbysample_withmem%>%
  left_join(more_relatedgen,by=c("GENUS","cluster"))%>%
  #if the ratio column is na, it means that the phylum-cluster key is not in the more_relatedgen, 
  #meaning that the genus is not that important in the cluster; thus exclude
  filter(!is.na(ratio))%>%
  select(-ratio)
sample_ofrelatedgen_bycluster
sample_ph<-selected.samples%>%select(ID,ph,envo_biome_2)%>%rename(SAMPLE=ID)
#combine sample_ofrelatedgen_bycluster with sample_ph
sorgc_ph<-sample_ofrelatedgen_bycluster%>%left_join(sample_ph)
sorgc_ph
```
## plot ph by genus composition
```{r}
new_labels<-c("1"="Weathered (acidic)","2"="Cold moist","3"="Dry (basic)","4"="Short vegatation\n(neutral)","5"="Dry cold","6"="Disturbed")
#facet grid is clearer than facet wrap
#take log for abundance to see clearer at the trend
p8<-ggplot(sorgc_ph,aes(x = ph,y=log(abundance),color=envo_biome_2))+
  geom_point()+
  facet_grid(cluster~GENUS,scales="free",labeller = labeller(cluster=new_labels,GENUS=label_wrap_gen(5)))+
  theme_bw()+
  xlab("pH")+ylab("genus composition")+
  theme(strip.text.x = element_text(8))
p8

```
## infer function of biomes
```{r}
otu_tax<-otus%>%
  select(-SEQUENCE)%>%
  mutate_all(na_if,"")%>%
  filter(!is.na(GENUS))%>%filter(GENUS!="g__")
otu_tax
tax<-unite(otu_tax,taxonomy,-1,sep = "; ",na.rm = TRUE)
tax
tax$taxonomy<-str_replace_all(tax$taxonomy,"\\[|\\]", "")
write.table(tax, file='otu_tax.tsv', quote=FALSE, sep='\t',row.names = FALSE)
tax<-tax%>%rename('seq'='taxonomy')

seq.func<-read_csv("seq_func.csv")%>%rename('func'='function')%>%rename('seq'='sequence')
seq.func
tax
taxfunc<-tax%>%left_join(seq.func)%>%select(ID,func)%>%distinct(ID,.keep_all=TRUE)
#map function with taxonomy path
taxfunc$func<-str_replace_all(taxfunc$func, '_', ' ')
long_func<-("methanogenesis by CO2 reduction with H2")
long_func2<-("aerobic chemoheterotrophy")
long_func3<-("photosynthetic cyanobacteria")
long_func4<-("fermentation")
taxfunc$func<-str_replace_all(taxfunc$func,long_func,"methano- genesis by CO2")
taxfunc$func<-str_replace_all(taxfunc$func,long_func2,"aerobic chemo- heterotrophy")
taxfunc$func<-str_replace_all(taxfunc$func,long_func3,"photo- synthetic cyanobacteria")
taxfunc$func<-str_replace_all(taxfunc$func,long_func4,"fermen- tation")
taxfunc
```
## plot community composition by function
```{r}
## community composition by fimctopm
## filter out no-genus data
community_composition_byfunc<-selected.otus3%>%
  left_join(community_composition_otu)%>%
  select(-ID)%>%
  group_by(func)%>%
  summarize_each(funs(sum))%>%
  filter(!is.na(func))
community_composition_byfunc
#add a column to highlight more prominant function
community_composit_tidy_func<-community_composition_byfunc%>%
  pivot_longer(!func,names_to="cluster",values_to="ratio")%>%
  mutate(ratio=100*ratio)%>%
  mutate(cluster=factor(cluster,levels=paste0("X",c(1:6))))%>%
  mutate(highlight=ifelse(ratio>0.5,"yes","no"))
community_composit_tidy_func
#prominent function within a cluster
pfwc<-community_composit_tidy_func[which(community_composit_tidy_func$ratio>0.55),]$func
pfwclevel<-levels(as.factor(pfwc))
pfwclevel
new_labels<-c("X1"="Weathered","X2"="Cold moist","X3"="Dry","X4"="Short vegetation","X5"="Cold dry","X6"="Disturbed")
p9<-ggplot(community_composit_tidy_func[which(community_composit_tidy_func$ratio>0.55),], aes(x=func,y=ratio)) +
  geom_col() +
  facet_wrap( ~ cluster, scales="free",ncol=3,labeller = labeller(cluster=new_labels)) +
  theme_bw()+
  aes(stringr::str_wrap(func, 2), ratio)+xlab("function")+
  theme(axis.text.x=element_text(size=rel(1.1)),axis.text.y = element_text(size=rel(0.8)))

p9.1<-ggplot(community_composit_tidy_func[community_composit_tidy_func$func%in%pfwclevel,], 
             aes(x=func,y=ratio,fill=highlight)) +
  geom_col() +
  facet_wrap( ~cluster,nrow=3,labeller = labeller(cluster=new_labels)) +
  theme_bw()+
  scale_fill_manual(values=c("yes"="tomato","no"="gray"),guide=FALSE)+
  aes(stringr::str_wrap(func, 2), ratio)+xlab("function")+
  theme(axis.text.x=element_text(size=rel(1.1)),axis.text.y = element_text(size=rel(0.8)))
p9
p9.1
ggsave("function ratio-cluster.jpg",p9.1,width = 1920/100, height=1080/100,dpi=600)
```
## map otu to function for each sample
```{r}
# join otu-sample with function
otufunc<-otus%>%select(ID)%>%left_join(taxfunc)%>%rename(OTU=ID)
otuSampleFunc<-read.table(infile, header=T, sep="\t", colClasses=c(rep("character", 2), "numeric"))%>%
  left_join(otufunc)
otuSampleFunc
#for each sample site, sum the otu number of each genus
#each sample site has 5000 otus
#exclude otus with unknown genus
perfuncsumbysample<-otuSampleFunc%>%
  group_by(SAMPLE,func)%>%
  summarize(amount=sum(VALUE))%>%
  mutate(abundance=(amount/5000)*100)%>%
  filter(!is.na(func))
#sample-clustermembership
sample_clustermem<-selected.samples%>%select(ID,cluster_mem)%>%rename(SAMPLE=ID)
#selected samples that have memberships(we only included samples that have otu num>9, that's why some samples were not included)
perfuncsumbysample_withmem<-perfuncsumbysample%>%
  filter(SAMPLE%in%sample_clustermem$SAMPLE)%>%
  left_join(sample_clustermem)%>%
  rename(cluster=cluster_mem)
perfuncsumbysample_withmem
#within each cluster, phyla that have more weight
more_relatedfunc<-(community_composit_tidy_func[which(community_composit_tidy_func$ratio>0.55),])%>%
  mutate_if(is.factor,as.character)%>%
  mutate(cluster=str_remove_all(cluster,"X"))%>%
  mutate(cluster=as.integer(cluster))
more_relatedfunc
perfuncsumbysample_withmem
#for each cluster, take the samples that have higher-weight-phyla in that cluster
sample_ofrelatedfunc_bycluster<-perfuncsumbysample_withmem%>%
  left_join(more_relatedfunc,by=c("func","cluster"))%>%
  #if the ratio column is na, it means that the phylum-cluster key is not in the more_relatedgen, 
  #meaning that the genus is not that important in the cluster; thus exclude
  filter(!is.na(ratio))%>%
  select(-ratio)
sample_ofrelatedfunc_bycluster
sample_ph<-selected.samples%>%select(ID,ph,envo_biome_2)%>%rename(SAMPLE=ID)
#combine sample_ofrelatedgen_bycluster with sample_ph
sorfc_ph<-sample_ofrelatedfunc_bycluster%>%left_join(sample_ph)
sorfc_ph
```
## plot ph by function composition
```{r}
new_labels<-c("1"="Weathered","2"="Cold moist","3"="Dry","4"="Short\nvegetation","5"="Cold dry","6"="Disturbed")
#facet grid is clearer than facet wrap
#take log for abundance to see clearer at the trend
p10<-ggplot(sorfc_ph,aes(x = ph,y=log10(abundance),color=envo_biome_2))+
  geom_point()+
  facet_grid(cluster~func,labeller = labeller(cluster=new_labels,func=label_wrap_gen(5)))+
  theme_bw()+
  xlab("pH")+ylab("log(function abundance)")+
  theme(strip.text= element_text(size=14),legend.position = "none")
p10
#ggplot default: dpi=30, height and width specified in 'in', ref: https://ggplot2.tidyverse.org/reference/ggsave.html#examples
#ref: 
ggsave("ph-function.jpg",p10,width=1920/100,height=1080/100,dpi=600)
```

## regression-function
```{r}
## function
sorfc_ph_reg<-sorfc_ph%>%mutate(logabundance=log10(abundance))%>%select(logabundance,cluster,func,ph)
sorfc_ph_reg
fit.func<-sorfc_ph_reg%>%group_by(cluster,func)%>%do(model=lm(logabundance~ph,data=.))

fit.funcdf<-lapply(fit.func$model,function(x){
  tidy(x)
})
#pvalues
pls<-lapply(fit.func$model,function(x){
  p<-tidy(x)[2,]$p.value
  return(p)
})%>%unlist
pls
estimatels<-lapply(fit.func$model, function(x){
  est<-tidy(x)[2,]$estimate
})%>%unlistls
lmsum.func<-data.frame(cbind(pls,estimatels))
fit.func.summ<-cbind(fit.func,lmsum.func)%>%rename(pval=pls,estimate=estimatels)%>%select(-model)
write_csv(fit.func.summ,file = "function_cluster_lmsum.csv")
#efficients that are significant
fit.func.sig<-fit.func.summ%>%filter(pval<0.05)
```
## regression-phyla
```{r}
## phyla
sorpc_ph

sorpc_ph_reg<-sorpc_ph%>%mutate(logabundance=log10(abundance))%>%select(logabundance,cluster,PHYLUM,ph)
sorpc_ph_reg
fit.func.phy<-sorpc_ph_reg%>%group_by(cluster,PHYLUM)%>%do(model=lm(logabundance~ph,data=.))

fit.funcdf.phy<-lapply(fit.func.phy$model,function(x){
  tidy(x)
})
fit.funcdf.phy
#pvalues
pls.phy<-lapply(fit.func.phy$model,function(x){
  p<-tidy(x)[2,]$p.value
  return(p)
})%>%unlist
pls.phy
estimatels.phy<-lapply(fit.func.phy$model, function(x){
  est<-tidy(x)[2,]$estimate
})%>%unlist
lmsum.func.phy<-data.frame(cbind(pls.phy,estimatels.phy))
fit.func.summ.phy<-cbind(fit.func.phy,lmsum.func.phy)%>%rename(pval=pls.phy,estimate=estimatels.phy)%>%select(-model)
write_csv(fit.func.summ.phy,file = "phyla_cluster_lmsum.csv")
#efficients that are significant
fit.func.sig.phy<-fit.func.summ.phy%>%filter(pval<0.05)
fit.func.sig.phy
```

